---
title: "Mood of the Week — Billboard Hot 100"
subtitle: "A population-vibe read from Spotify audio features in this week’s Hot 100"
format:
  dashboard:
    logo: "images/logo.png"
    orientation: columns
---

```{=html}
<div class="d-flex flex-wrap gap-2 mb-2">
  <a class="btn btn-outline-light btn-sm rounded-pill px-3" href="about.qmd">ℹ️ About Me</a>
  <a class="btn btn-outline-light btn-sm rounded-pill px-3" href="dashboard.qmd">📊 Hot 100 Chart</a>
  <a class="btn btn-success btn-sm rounded-pill px-3" href="mood-of-the-week.qmd">🎧 Mood Snapshot</a>
</div>
```

A quick read on this week’s chart atmosphere using Spotify audio features. **Mood Index** is a composite score from valence, energy, danceability, and tempo (scaled to 0–100), shown as both track-average and popularity-weighted exposure.

```{r}
#| include: false
library(tidyverse)

hot100_mood <- readRDS("data/hot100_enriched.rds")

if (!"song_id" %in% names(hot100_mood) && "reccobeats_id" %in% names(hot100_mood)) {
  hot100_mood <- hot100_mood |> mutate(song_id = as.character(reccobeats_id))
}

if (!"weeks_on_chart" %in% names(hot100_mood)) {
  hot100_mood <- hot100_mood |> mutate(weeks_on_chart = 1L)
}

hot100_mood <- hot100_mood |>
  mutate(
    song_id = as.character(song_id),
    popularity = as.numeric(popularity),
    valence = as.numeric(valence),
    energy = as.numeric(energy),
    danceability = as.numeric(danceability),
    tempo = as.numeric(tempo),
    weeks_on_chart = as.integer(weeks_on_chart)
  ) |>
  drop_na(song_id, song_name, artist_name, valence, energy, danceability, tempo)

weighted_mean_safe <- function(x, w) {
  ok <- is.finite(x) & is.finite(w)
  if (!any(ok)) return(NA_real_)
  weighted.mean(x[ok], w[ok])
}

tempo_score <- function(x) pmax(pmin((x - 60) / 120, 1), 0)

mood_index <- function(df, weighted = FALSE) {
  if (!nrow(df)) return(NA_real_)
  if (weighted) {
    v <- weighted_mean_safe(df$valence, df$popularity)
    e <- weighted_mean_safe(df$energy, df$popularity)
    d <- weighted_mean_safe(df$danceability, df$popularity)
    t <- weighted_mean_safe(tempo_score(df$tempo), df$popularity)
  } else {
    v <- mean(df$valence, na.rm = TRUE)
    e <- mean(df$energy, na.rm = TRUE)
    d <- mean(df$danceability, na.rm = TRUE)
    t <- mean(tempo_score(df$tempo), na.rm = TRUE)
  }
  100 * (0.35 * v + 0.35 * e + 0.2 * d + 0.1 * t)
}

med_v <- median(hot100_mood$valence, na.rm = TRUE)
med_e <- median(hot100_mood$energy, na.rm = TRUE)

hot100_mood <- hot100_mood |>
  mutate(
    archetype = case_when(
      valence >= med_v & energy >= med_e ~ "Bright & Driven",
      valence >= med_v & energy < med_e ~ "Warm & Calm",
      valence < med_v & energy >= med_e ~ "Tense & Urgent",
      TRUE ~ "Melancholic & Low-Arousal"
    )
  )

kpi <- hot100_mood |>
  summarise(
    mood_unweighted = mood_index(cur_data(), FALSE),
    mood_weighted = mood_index(cur_data(), TRUE),
    valence_unweighted = mean(valence, na.rm = TRUE),
    valence_weighted = weighted_mean_safe(valence, popularity),
    energy_unweighted = mean(energy, na.rm = TRUE),
    energy_weighted = weighted_mean_safe(energy, popularity),
    tempo_median = median(tempo, na.rm = TRUE)
  )

quad_summary <- hot100_mood |>
  count(archetype, name = "tracks") |>
  left_join(
    hot100_mood |> group_by(archetype) |> summarise(pop_sum = sum(popularity, na.rm = TRUE), .groups = "drop"),
    by = "archetype"
  ) |>
  mutate(
    track_share = tracks / sum(tracks),
    weighted_share = pop_sum / sum(pop_sum)
  )

happiest <- hot100_mood |> arrange(desc(valence)) |> slice_head(n = 5)
energetic <- hot100_mood |> arrange(desc(energy)) |> slice_head(n = 5)
melancholic <- hot100_mood |> mutate(low_combo = valence + energy) |> arrange(low_combo) |> slice_head(n = 5)

mood_palette <- c(
  "Bright & Driven" = "#1DB954",
  "Warm & Calm" = "#EDDE11",
  "Tense & Urgent" = "#ED8E11",
  "Melancholic & Low-Arousal" = "#7D3C98"
)

card_style <- "background:#282828;border:1px solid #3b3b3b;border-radius:10px;padding:12px;height:100%;"
```

## Column {width="100%"}

### Row {height="17%"}

```{r}
#| echo: false
#| results: asis
cat(
  "<div class='row g-2'>",
  sprintf("<div class='col-3'><div style='%s'><h4>Mood Index</h4><p style='margin-bottom:0;'>%.1f (avg) / %.1f (weighted)</p></div></div>", card_style, kpi$mood_unweighted, kpi$mood_weighted),
  sprintf("<div class='col-3'><div style='%s'><h4>Avg Valence</h4><p style='margin-bottom:0;'>%.3f / %.3f</p></div></div>", card_style, kpi$valence_unweighted, kpi$valence_weighted),
  sprintf("<div class='col-3'><div style='%s'><h4>Avg Energy</h4><p style='margin-bottom:0;'>%.3f / %.3f</p></div></div>", card_style, kpi$energy_unweighted, kpi$energy_weighted),
  sprintf("<div class='col-3'><div style='%s'><h4>Median Tempo</h4><p style='margin-bottom:0;'>%.0f BPM</p></div></div>", card_style, kpi$tempo_median),
  "</div>"
)
```

### Row {height="45%"}

#### Energy × Valence (plain-language mood quadrants)

```{r}
#| echo: false
corner_lab <- quad_summary |>
  arrange(archetype) |>
  mutate(label = sprintf("%s: %.1f%% / w %.1f%%", archetype, 100 * track_share, 100 * weighted_share)) |>
  pull(label) |>
  paste(collapse = " • ")

cat(sprintf("<p style='color:#d9d9d9;font-size:0.95rem;margin-bottom:6px;'>%s</p>", corner_lab))

plot(
  ggplot(hot100_mood, aes(x = valence, y = energy)) +
    geom_hline(yintercept = med_e, linetype = "dashed", color = "#f4f4f4") +
    geom_vline(xintercept = med_v, linetype = "dashed", color = "#f4f4f4") +
    geom_point(aes(color = archetype, size = popularity), alpha = 0.8) +
    scale_color_manual(values = mood_palette) +
    scale_size(range = c(2.5, 10)) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "Valence", y = "Energy", color = "Mood archetype", size = "Popularity") +
    theme_minimal(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "#121212", color = NA),
      plot.background = element_rect(fill = "#121212", color = NA),
      legend.background = element_rect(fill = "#121212", color = NA),
      legend.key = element_rect(fill = "#121212", color = NA),
      text = element_text(color = "white"),
      axis.text = element_text(color = "white"),
      axis.title = element_text(color = "white"),
      legend.title = element_text(color = "white"),
      legend.text = element_text(color = "white"),
      panel.grid.major = element_line(color = "#282828"),
      panel.grid.minor = element_blank()
    )
)
```

### Row {height="38%"}

```{r}
#| echo: false
plot_data <- hot100_mood |>
  select(valence, energy) |>
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value")

feature_lines <- tibble(
  feature = c("valence", "valence", "energy", "energy"),
  line_type = c("Average", "Weighted average", "Average", "Weighted average"),
  xintercept = c(
    mean(hot100_mood$valence, na.rm = TRUE),
    weighted_mean_safe(hot100_mood$valence, hot100_mood$popularity),
    mean(hot100_mood$energy, na.rm = TRUE),
    weighted_mean_safe(hot100_mood$energy, hot100_mood$popularity)
  )
)

plot(
  ggplot(plot_data, aes(x = value)) +
    geom_histogram(fill = "#1DB954", color = "#121212", bins = 22, alpha = 0.75) +
    geom_vline(data = feature_lines, aes(xintercept = xintercept, linetype = line_type), color = "#f4f4f4", linewidth = 0.7) +
    facet_wrap(~feature, ncol = 2, scales = "free_y", labeller = as_labeller(c(valence = "Valence distribution", energy = "Energy distribution"))) +
    scale_linetype_manual(values = c("Average" = "dashed", "Weighted average" = "solid")) +
    labs(x = NULL, y = "Tracks", linetype = NULL) +
    theme_minimal(base_size = 12) +
    theme(
      panel.background = element_rect(fill = "#121212", color = NA),
      plot.background = element_rect(fill = "#121212", color = NA),
      legend.background = element_rect(fill = "#121212", color = NA),
      legend.key = element_rect(fill = "#121212", color = NA),
      text = element_text(color = "white"),
      axis.text = element_text(color = "white"),
      axis.title = element_text(color = "white"),
      strip.background = element_rect(fill = "#282828", color = NA),
      strip.text = element_text(color = "white"),
      legend.text = element_text(color = "white")
    )
)
```

```{r}
#| echo: false
comp_counts <- hot100_mood |> count(archetype, name = "value")

comp_weighted <- hot100_mood |>
  group_by(archetype) |>
  summarise(value = sum(popularity, na.rm = TRUE), .groups = "drop") |>
  mutate(value = value / sum(value))

p_counts <- ggplot(comp_counts, aes(x = archetype, y = value, fill = archetype)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = mood_palette) +
  labs(x = NULL, y = "Tracks", title = "Track count") +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "#121212", color = NA),
    plot.background = element_rect(fill = "#121212", color = NA),
    legend.position = "none",
    text = element_text(color = "white"),
    axis.text = element_text(color = "white", size = 8),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", size = 10),
    panel.grid.minor = element_blank()
  )

p_weighted <- ggplot(comp_weighted, aes(x = archetype, y = value, fill = archetype)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = mood_palette) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "Share", title = "Popularity-weighted share") +
  theme_minimal(base_size = 11) +
  theme(
    panel.background = element_rect(fill = "#121212", color = NA),
    plot.background = element_rect(fill = "#121212", color = NA),
    legend.position = "none",
    text = element_text(color = "white"),
    axis.text = element_text(color = "white", size = 8),
    axis.title = element_text(color = "white"),
    plot.title = element_text(color = "white", size = 10),
    panel.grid.minor = element_blank()
  )

cat("<h4>Mood archetype composition</h4>")
plot(p_counts)
plot(p_weighted)
```

## Column {width="100%"}

### Row {height="27%"}

```{r}
#| echo: false
#| results: asis
make_outlier_rows <- function(df) {
  pmap_chr(df |> select(album_art, song_name, artist_name, weeks_on_chart), function(album_art, song_name, artist_name, weeks_on_chart) {
    sprintf(
      "<div style='display:flex;gap:10px;align-items:center;margin:0 0 10px 0;'><img src='%s' style='width:54px;height:54px;border-radius:8px;object-fit:cover;flex:0 0 54px;'><div style='line-height:1.25;'><div style='font-weight:600;'>%s</div><div style='font-size:.9em;color:#d8d8d8;'>%s</div><div style='font-size:.85em;color:#d8d8d8;'><em>weeks_on_chart</em>: %s</div></div></div>",
      album_art, song_name, artist_name, weeks_on_chart
    )
  })
}

cat("<div class='row g-3'>")
cat("<div class='col-4'><div style='", card_style, "'><h4>Happiest</h4>", paste(make_outlier_rows(happiest), collapse = ""), "</div></div>")
cat("<div class='col-4'><div style='", card_style, "'><h4>Most Energetic</h4>", paste(make_outlier_rows(energetic), collapse = ""), "</div></div>")
cat("<div class='col-4'><div style='", card_style, "'><h4>Most Melancholic / Low-Arousal</h4>", paste(make_outlier_rows(melancholic), collapse = ""), "</div></div>")
cat("</div>")
```

### Row {height="18%"}

Mood is inferred from Spotify audio features, so this is an aggregate signal of track acoustics rather than direct listener emotion. Popularity weighting is a proxy for exposure, shaping this population-vibe read.

*Tracking started March 15, 2026; values reflect weeks observed since that date.*

### Row {height="16%"}

#### Optional song preview

```{ojs}
hot100 = await FileAttachment("data/hot100_ojs.json").json()

viewof songDropdown = Inputs.select(
  hot100,
  {
    label: "Choose a song",
    format: d => `${d.song_name} — ${d.artist_name}`,
    unique: true
  }
)
```

```{ojs}
html`<iframe class="custom-iframe" style="border-radius:12px; display:flex; justify-content:center; align-items:center;" 
  src=${`https://open.spotify.com/embed/track/${songDropdown.song_id}?utm_source=generator&theme=0`} 
  width="100%" 
  height="180" 
  frameBorder="0" 
  allowfullscreen="" 
  allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture" 
  loading="lazy">
</iframe>`
```

```{=html}
<div class="d-flex flex-wrap gap-2 mt-3">
  <a class="btn btn-outline-light btn-sm rounded-pill px-3" href="about.qmd">ℹ️ About Me</a>
  <a class="btn btn-outline-light btn-sm rounded-pill px-3" href="dashboard.qmd">📊 Hot 100 Chart</a>
  <a class="btn btn-success btn-sm rounded-pill px-3" href="mood-of-the-week.qmd">🎧 Mood Snapshot</a>
</div>
```
