---
title: "Mood of the Week — Billboard Hot 100"
subtitle: "A population-vibe read from Spotify audio features in this week's Hot 100"
format:
  html:
    page-layout: full
    theme: darkly
    css: mood-of-the-week.css
    toc: false
    self-contained: false
---

```{r}
#| include: false
library(tidyverse)
library(scales)

hot100_mood <- readRDS("data/hot100_enriched.rds")

if (!"song_id" %in% names(hot100_mood) && "reccobeats_id" %in% names(hot100_mood)) {
  hot100_mood <- hot100_mood |> mutate(song_id = as.character(reccobeats_id))
}
if (!"weeks_on_chart" %in% names(hot100_mood)) {
  hot100_mood <- hot100_mood |> mutate(weeks_on_chart = 1L)
}

hot100_mood <- hot100_mood |>
  mutate(
    song_id        = as.character(song_id),
    popularity     = as.numeric(popularity),
    valence        = as.numeric(valence),
    energy         = as.numeric(energy),
    danceability   = as.numeric(danceability),
    tempo          = as.numeric(tempo),
    weeks_on_chart = as.integer(weeks_on_chart)
  ) |>
  drop_na(song_id, song_name, artist_name, valence, energy, danceability, tempo)

weighted_mean_safe <- function(x, w) {
  ok <- is.finite(x) & is.finite(w)
  if (!any(ok)) return(NA_real_)
  weighted.mean(x[ok], w[ok])
}

tempo_score <- function(x) pmax(pmin((x - 60) / 120, 1), 0)

mood_index <- function(df, weighted = FALSE) {
  if (!nrow(df)) return(NA_real_)
  if (weighted) {
    v <- weighted_mean_safe(df$valence,            df$popularity)
    e <- weighted_mean_safe(df$energy,             df$popularity)
    d <- weighted_mean_safe(df$danceability,       df$popularity)
    t <- weighted_mean_safe(tempo_score(df$tempo), df$popularity)
  } else {
    v <- mean(df$valence,            na.rm = TRUE)
    e <- mean(df$energy,             na.rm = TRUE)
    d <- mean(df$danceability,       na.rm = TRUE)
    t <- mean(tempo_score(df$tempo), na.rm = TRUE)
  }
  100 * (0.35 * v + 0.35 * e + 0.2 * d + 0.1 * t)
}

med_v <- median(hot100_mood$valence, na.rm = TRUE)
med_e <- median(hot100_mood$energy,  na.rm = TRUE)

hot100_mood <- hot100_mood |>
  mutate(
    archetype = case_when(
      valence >= med_v & energy >= med_e ~ "Bright & Driven",
      valence >= med_v & energy <  med_e ~ "Warm & Calm",
      valence <  med_v & energy >= med_e ~ "Tense & Urgent",
      TRUE                               ~ "Melancholic & Low-Arousal"
    )
  )

kpi <- hot100_mood |>
  summarise(
    mood_unweighted    = mood_index(cur_data(), FALSE),
    mood_weighted      = mood_index(cur_data(), TRUE),
    valence_unweighted = mean(valence, na.rm = TRUE),
    valence_weighted   = weighted_mean_safe(valence, popularity),
    energy_unweighted  = mean(energy,  na.rm = TRUE),
    energy_weighted    = weighted_mean_safe(energy,  popularity),
    tempo_median       = median(tempo, na.rm = TRUE)
  )

quad_summary <- hot100_mood |>
  count(archetype, name = "tracks") |>
  left_join(
    hot100_mood |>
      group_by(archetype) |>
      summarise(pop_sum = sum(popularity, na.rm = TRUE), .groups = "drop"),
    by = "archetype"
  ) |>
  mutate(
    track_share    = tracks / sum(tracks),
    weighted_share = pop_sum / sum(pop_sum)
  )

corner_lab <- quad_summary |>
  arrange(archetype) |>
  mutate(label = sprintf("%s: %.1f%% / w %.1f%%", archetype, 100 * track_share, 100 * weighted_share)) |>
  pull(label) |>
  paste(collapse = "  •  ")

happiest    <- hot100_mood |> arrange(desc(valence))                                     |> slice_head(n = 5)
energetic   <- hot100_mood |> arrange(desc(energy))                                      |> slice_head(n = 5)
melancholic <- hot100_mood |> mutate(low_combo = valence + energy) |> arrange(low_combo) |> slice_head(n = 5)

mood_palette <- c(
  "Bright & Driven"           = "#1DB954",
  "Warm & Calm"               = "#EDDE11",
  "Tense & Urgent"            = "#ED8E11",
  "Melancholic & Low-Arousal" = "#7D3C98"
)

dark_theme <- theme_minimal(base_size = 13) +
  theme(
    panel.background  = element_rect(fill = "#121212", color = NA),
    plot.background   = element_rect(fill = "#121212", color = NA),
    legend.background = element_rect(fill = "#121212", color = NA),
    legend.key        = element_rect(fill = "#121212", color = NA),
    text              = element_text(color = "#ffffff", family = "Montserrat"),
    axis.text         = element_text(color = "#cccccc"),
    axis.title        = element_text(color = "#ffffff"),
    legend.title      = element_text(color = "#ffffff"),
    legend.text       = element_text(color = "#ffffff"),
    plot.subtitle     = element_text(color = "#aaaaaa", size = 9),
    plot.title        = element_text(color = "#ffffff"),
    panel.grid.major  = element_line(color = "#2a2a2a"),
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "#1e1e1e", color = NA),
    strip.text        = element_text(color = "#ffffff", face = "bold")
  )
```

::: {.mood-page}

::: {.mood-nav}
<a class="mood-nav-btn" href="about.html">ℹ️ About Me</a>
<a class="mood-nav-btn" href="dashboard.html">📊 Hot 100 Chart</a>
<a class="mood-nav-btn mood-nav-btn-active" href="mood-of-the-week.html">🎧 Mood Snapshot</a>
:::

::: {.mood-intro}
**Mood Index** is a composite score from valence, energy, danceability, and tempo (scaled 0–100), shown as both track-average and popularity-weighted exposure.
:::

::: {.mood-section-header}
KPIs
:::

```{r}
#| echo: false
#| results: asis
kpis <- list(
  list(label = "Mood Index",   value = sprintf("%.1f", kpi$mood_unweighted),    sub = sprintf("%.1f weighted", kpi$mood_weighted)),
  list(label = "Avg Valence",  value = sprintf("%.3f", kpi$valence_unweighted), sub = sprintf("%.3f weighted", kpi$valence_weighted)),
  list(label = "Avg Energy",   value = sprintf("%.3f", kpi$energy_unweighted),  sub = sprintf("%.3f weighted", kpi$energy_weighted)),
  list(label = "Median Tempo", value = sprintf("%.0f BPM", kpi$tempo_median),   sub = "&nbsp;")
)

cat('<div class="mood-kpi-row">')
for (k in kpis) {
  cat(sprintf(
    '<div class="mood-kpi-card">\n       <div class="mood-kpi-label">%s</div>\n       <div class="mood-kpi-value">%s</div>\n       <div class="mood-kpi-sub">%s</div>\n     </div>',
    k$label, k$value, k$sub
  ))
}
cat('</div>')
```

::: {.mood-section-header}
Energy × Valence
:::

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6
#| out-width: 100%
ggplot(hot100_mood, aes(x = valence, y = energy)) +
  geom_hline(yintercept = med_e, linetype = "dashed", color = "#555555", linewidth = 0.5) +
  geom_vline(xintercept = med_v, linetype = "dashed", color = "#555555", linewidth = 0.5) +
  geom_point(aes(color = archetype, size = popularity), alpha = 0.85) +
  scale_color_manual(values = mood_palette) +
  scale_size(range = c(3, 11)) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x        = "Valence  →  more positive",
    y        = "Energy  →  more intense",
    color    = "Mood archetype",
    size     = "Popularity",
    subtitle = corner_lab
  ) +
  dark_theme
```

::: {.mood-section-header}
Distributions
:::

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4
#| out-width: 100%
plot_data <- hot100_mood |>
  select(valence, energy) |>
  pivot_longer(cols = everything(), names_to = "feature", values_to = "value")

feature_lines <- tibble(
  feature    = c("valence", "valence", "energy", "energy"),
  line_type  = c("Average", "Weighted avg", "Average", "Weighted avg"),
  xintercept = c(
    mean(hot100_mood$valence, na.rm = TRUE),
    weighted_mean_safe(hot100_mood$valence, hot100_mood$popularity),
    mean(hot100_mood$energy,  na.rm = TRUE),
    weighted_mean_safe(hot100_mood$energy,  hot100_mood$popularity)
  )
)

ggplot(plot_data, aes(x = value)) +
  geom_histogram(fill = "#1DB954", color = "#121212", bins = 25, alpha = 0.8) +
  geom_vline(
    data      = feature_lines,
    aes(xintercept = xintercept, linetype = line_type),
    color     = "#ffffff",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~feature, ncol = 2, scales = "free_y",
    labeller = as_labeller(c(valence = "Valence distribution", energy = "Energy distribution"))
  ) +
  scale_linetype_manual(values = c("Average" = "dashed", "Weighted avg" = "solid")) +
  labs(x = NULL, y = "Tracks", linetype = NULL) +
  dark_theme
```

::: {.mood-section-header}
Archetype Composition
:::

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 4
#| out-width: 100%
comp_counts <- hot100_mood |>
  count(archetype, name = "value") |>
  mutate(metric = "Track count")

comp_weighted <- hot100_mood |>
  group_by(archetype) |>
  summarise(raw = sum(popularity, na.rm = TRUE), .groups = "drop") |>
  mutate(value = raw / sum(raw), metric = "Popularity-weighted share")

comp_all <- bind_rows(
  comp_counts   |> select(archetype, value, metric),
  comp_weighted |> select(archetype, value, metric)
) |>
  mutate(metric = factor(metric, levels = c("Track count", "Popularity-weighted share")))

ggplot(comp_all, aes(x = reorder(archetype, -value), y = value, fill = archetype)) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = mood_palette) +
  facet_wrap(~metric, scales = "free_y") +
  scale_y_continuous(labels = function(x) ifelse(x <= 1, percent(x, accuracy = 1), as.character(round(x)))) +
  labs(x = NULL, y = NULL) +
  dark_theme +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(size = 9, angle = 12, hjust = 1)
  )
```

::: {.mood-section-header}
Top Tracks
:::

```{r}
#| echo: false
#| results: asis
make_cards <- function(df, title) {
  rows <- pmap_chr(
    df |> select(album_art, song_name, artist_name, weeks_on_chart),
    function(album_art, song_name, artist_name, weeks_on_chart) {
      sprintf(
        '<div class="mood-track-item">\n           <img src="%s" class="mood-track-img">\n           <div class="mood-track-info">\n             <div class="mood-track-name">%s</div>\n             <div class="mood-track-artist">%s</div>\n             <div class="mood-track-weeks">%s wks on chart</div>\n           </div>\n         </div>',
        album_art, song_name, artist_name, weeks_on_chart
      )
    }
  )
  sprintf(
    '<div class="mood-outlier-card">\n       <div class="mood-outlier-title">%s</div>\n       %s\n     </div>',
    title, paste(rows, collapse = "")
  )
}

cat('<div class="mood-outlier-row">')
cat(make_cards(happiest,    "😊 Happiest"))
cat(make_cards(energetic,   "⚡ Most Energetic"))
cat(make_cards(melancholic, "🌑 Most Melancholic"))
cat('</div>')
```

::: {.mood-section-header}
Song Preview
:::

```{ojs}
hot100 = await FileAttachment("data/hot100_ojs.json").json()

viewof songDropdown = Inputs.select(
  hot100,
  {
    label: "Choose a song",
    format: d => `${d.song_name} — ${d.artist_name}`,
    unique: true
  }
)
```

```{ojs}
html`<iframe
  style="border-radius:12px;margin-top:10px;display:block;"
  src=${`https://open.spotify.com/embed/track/${songDropdown.song_id}?utm_source=generator&theme=0`}
  width="100%"
  height="180"
  frameBorder="0"
  allowfullscreen=""
  allow="autoplay; clipboard-write; encrypted-media; fullscreen; picture-in-picture"
  loading="lazy">
</iframe>`
```

::: {.mood-footer}
Mood is inferred from Spotify audio features — an aggregate signal of track acoustics, not direct listener emotion. Popularity weighting proxies exposure. *Tracking started March 15, 2026.*
:::

::: {.mood-nav}
<a class="mood-nav-btn" href="about.html">ℹ️ About Me</a>
<a class="mood-nav-btn" href="dashboard.html">📊 Hot 100 Chart</a>
<a class="mood-nav-btn mood-nav-btn-active" href="mood-of-the-week.html">🎧 Mood Snapshot</a>
:::

:::
